---
title: "GeneSet: Representing Gene Sets in the Tidyverse"
author: "Kayla Morrell"
date: "`r Sys.Date()`"
output:
    BiocStyle::html_document
package: GeneSet
vignette: >
    %\VignetteIndexEntry{GeneSet: Representing Gene Sets in the Tidyverse}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEndcoding{UTF-8}
---

# Introduction
`GeneSet` is a package that represents gene sets in a tibble format with the `tbl_geneset` class. Gene sets are read in and converted into a tibble format. From here, typical `dplyr` operations can be performed on the tibble gene set.

# Installation
Install the most recent version of Bioconductor:

```{r bioconductor, eval = FALSE}
if(!requireNamespace("BiocManager", quietly = TRUE))
    install.pacakges("BiocManager")
BiocManager::install("GeneSet")
```

The development version is also available for install from GitHub:

```{r github, eval = FALSE}
BiocManager::install("Kayla-Morrell/GeneSet")
```

Then load `GeneSet`:

```{r  load, message = FALSE}
library(GeneSet)
```
# GeneSet

## Input and Output

`GeneSet` can create a `tbl_geneset` using two different input methods. The first is to input named character vectors of gene sets. `tbl_geneset()` returns the vectors as `tbl_geneset` tibble containing two columns, `gene` and `set`.

```{r constructor}
tbl <- tbl_geneset(set1 = letters, set2 = LETTERS)
tbl
```

The second method of creating a `tbl_geneset` would be to read in `.gmt` files. Using `import()` a path to a downloaded `.gmt` file is read in and a `tbl_geneset` is returned. The example below uses a hallmark gene set downloaded from [GSEA][], which is also included with this package. This `tbl_geneset` also includes a `source` column for reference as to where the gene set came from.

[GSEA]: http://software.broadinstitute.org/gsea/index.jsp

```{r gmt}
gmtFile <- system.file(package = "GeneSet", "extdata", "hallmark.gene.symbol.gmt")
tbl2 <- import(gmtFile)
tbl2
```

<!-- example of export -->
`export()` allows for a `tbl_geneset` to be exported into a temporary file with the extention `.gmt`.

```{r export}
fl <- tempfile(fileext = ".gmt")
gmt <- export(tbl2, fl)
gmt
```
<!-- Advanced use (construct gmt file) -->

## Implemented functions (or functionality)

`GeneSet` adopts the use of many `dplyr` functions such as `filter()`, `select()`, `mutate()`, `group_by()`, `ungroup()`, `summarize()`, and `arrange()`.

```{r install, message = FALSE}
library(dplyr)
```

```{r examples}
tbl <- tbl_geneset(set1 = letters, set2 = LETTERS)
tbl
tbl %>% filter(gene == "a" | gene == "A")
tbl %>% mutate(pval = rnorm(dim(tbl)[1]))
tbl %>% group_by(set) %>% summarise(n = n())
```

# Case study

<!-- airway package, dataset called airway (DESeq2 package vignette) -->

<!-- mapping ensemble id to GO id or KEGG id, creating gene set for relevant pathways -->

<!-- subsetting airway dataset to contain only genes in relevant pathways -->

# Session info

```{r}
sessionInfo()
```