---
title: "GeneSet: Representing Gene Sets in the Tidyverse"
author: "Kayla Morrell"
date: "`r Sys.Date()`"
output:
    BiocStyle::html_document:
	toc: true
	toc_depth: 2
vignette: >
  %\VignetteIndexEntry{GeneSet: Representing Gene Sets in the Tidyverse}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction
`GeneSet` is a package that represents gene sets in a tibble format with the `tbl_geneset` class. Gene sets are read in and converted into a tibble format. From here, typical `dplyr` operations can be performed on the tibble gene set. 

# Installation

Install the most recent version of Bioconductor:

```{r bioconductor, eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
   install.packages("BiocManager")
BiocManager::install("GeneSet")
```

The development version is also available for install from GitHub:

``` {r github, eval = FALSE}
BiocManager::install("Kayla-Morrell/GeneSet")
```

Then load `GeneSet`:

``` {r package, results = 'hide', message = FALSE}
library(GeneSet)
```

# GeneSet

## Input and Output
GeneSet
