
---
title: "GeneSet: Representing Gene Sets in the Tidyverse"
author: "Kayla Morrell"
date: "`r Sys.Date()`"
output:
    BiocStyle::html_document
package: GeneSet
vignette: >
    %\VignetteIndexEntry{GeneSet: Representing Gene Sets in the Tidyverse}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEndcoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

# Introduction
`GeneSet` is a package that represents gene sets in a tibble format with the `tbl_geneset` class. Gene sets are read in and converted into a tibble format. From here, typical `dplyr` operations can be performed on the tibble gene set.

# Installation
Install the most recent version of Bioconductor:

```{r bioconductor, eval = FALSE}
if(!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("GeneSet")
```

The development version is also available for install from GitHub:

```{r github, eval = FALSE}
BiocManager::install("Kayla-Morrell/GeneSet")
```

Then load `GeneSet`:

```{r  load, message = FALSE}
library(GeneSet)
```
# GeneSet

## Input and Output

`GeneSet` can create a `tbl_geneset` using two different input methods. The first is to input named character vectors of gene sets. `tbl_geneset()` returns the vectors as `tbl_geneset` tibble containing two columns, `gene` and `set`.

```{r constructor}
tbl <- tbl_geneset(set1 = letters, set2 = LETTERS)
tbl
```

The second method of creating a `tbl_geneset` would be to read in `.gmt` files. Using `import()`, a path to a downloaded `.gmt` file is read in and a `tbl_geneset` is returned. The example below uses a hallmark gene set downloaded from [GSEA][], which is also included with this package. This `tbl_geneset` includes a `source` column for reference as to where the gene set came from.

[GSEA]: http://software.broadinstitute.org/gsea/index.jsp

```{r gmt}
gmtFile <- system.file(package = "GeneSet", "extdata",
	   		       "hallmark.gene.symbol.gmt")
tbl2 <- import(gmtFile)
tbl2
```

`export()` allows for a `tbl_geneset` to be exported into a temporary file with the extention `.gmt`.

```{r export}
fl <- tempfile(fileext = ".gmt")
gmt <- export(tbl2, fl)
gmt
```

## Implemented functions

`GeneSet` adopts the use of many `dplyr` functions such as `filter()`, `select()`, `mutate()`, `group_by()`, `ungroup()`, `summarize()`, and `arrange()`.

```{r install, message = FALSE}
library(dplyr)
```

```{r examples}
tbl <- tbl_geneset(set1 = letters, set2 = LETTERS)
tbl
tbl %>% filter(gene == "a" | gene == "A")
tbl %>% mutate(pval = rnorm(dim(tbl)[1]))
tbl %>% group_by(set) %>% summarise(n = n())
```

# Case study
Next, we demonstrate the use of `GeneSet` with an experiement dataset `airway` from the package `airway`. This data is from an RNA-Seq experiment on airway smooth muscle (ASM) cell lines.

The first step is to load the library and the necessary data.

```{r airway, message = FALSE}
library("airway")
data("airway")
se <- airway
```

The pacakges `AnnotationDbi` and `org.Hs.eg.db` are used for the mapping of gene symbols, GO ids, and Entrez ids to the already established Ensembl ids in `airway`. These Ensembl ids are the rownames of `airway`.

```{r mapping, message = FALSE}
library("AnnotationDbi")
library("org.Hs.eg.db")

rowData(se)$symbol <- mapIds(org.Hs.eg.db, keys = rownames(se),
		      			   column = "SYMBOL", keytype = "ENSEMBL",
					   multiVals = "first")

rowData(se)$go <- mapIds(org.Hs.eg.db, keys = rownames(se),
				       column = "GO", keytype = "ENSEMBL",
				       multiVals = "first")

rowData(se)$entrez <- mapIds(org.Hs.eg.db, keys = rownames(se),
			    		   column = "ENTREZID", keytype = "ENSEMBL",
					   multiVals = "first")

rowData(se)
```

The library `KEGGREST` is a client interface to the KEGG REST server. KEGG contains pathway maps that represent interaction, reaction and relation networks for various biological processes and diseases. Below we use `KEGGREST` to develop a `tbl_geneset` that contains the genes for every pathway map in KEGG.

```{r kegg, message = FALSE}
library(KEGGREST)
paths <- keggList("pathway", "hsa")
codes <- sub("path:", "", names(paths))
genes <- sapply(codes, function(x) {
      	 	       	path <- keggGet(x)
			     path[[1]]$GENE[c(TRUE, FALSE)]
			})
genes <- genes[lengths(genes) != 0]

tbl <- do.call(tbl_geneset, genes)
tbl <- tbl %>% mutate(ensembl = mapIds(org.Hs.eg.db, keys = tbl$gene,
				    		     column = "ENSEMBL", keytype = "ENTREZID",
				   		      multivals = "first"))

tbl
```
Since we are working with ASM data we thought we would subset the `airway` data to contain only the genes in the asthma pathway. Asthma has the pathway id of `hsa05310` and contains 31 genes.

```{r subset}
asthma <- tbl %>% filter(set == "hsa05310")

se <- se[rownames(se) %in% asthma$ensembl,]

se
rowData(se)
```

This filtering can also be done for multiple pathways. 

```{r multiple}
pathways <- c("hsa05310", "hsa04110", "hsa05224", "hsa04970")
multipaths <- filter(tbl, set %in% pathways)

multipaths
```

# Session info

```{r}
sessionInfo()
```