---
title: "GeneSet: Representing Gene Sets in the Tidyverse"
author: "Kayla Morrell"
date: "`r Sys.Date()`"
output:
    BiocStyle::html_document:
	toc: true
	toc_depth: 2
vignette: >
    %\VignetteIndexEntry{GeneSet: Representing Gene Sets in the Tidyverse}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEndcoding{UTF-8}
---

#Introduction
`GeneSet` is a package that represents gene sets in a tibble format with the `tbl_geneset` class. Gene sets are read in and converted into a tibble format. From here, typical `dplyr` operations can be performed on the tibble gene set.

#Installation
Install the most recent version of Bioconductor:

```{r bioconductor, eval = FALSE}
if(!requireNamespace("BiocManager", quietly = TRUE))
    install.pacakges("BiocManager")
BiocManager::install("GeneSet")
```

The development version is also available for install from GitHub:

```{r github, eval = FALSE}
BiocManager::install("Kayla-Morrell/GeneSet")
```

Then load `GeneSet`:

```{r  load, message = FALSE}
library(GeneSet)
```
#GeneSet

##Input and Output